\documentclass[10pt]{article}
\input{common}
\title{Adaptation Experiment 1: Univariate Control}
\author{}
\date{}
\begin{document}
\maketitle

\section{Theoretical overview and goal}
A forecasting adapter is useful only if it improves a well-defined, untouched
backbone.  For a look-back $x\in\mathbb{R}^{L}$, a pretrained forecaster produces
$\hat y_0=f(x)\in\mathbb{R}^{H}$.  The univariate control evaluates $f$ directly,
without retrieved neighbors, gates, or trainable adaptation.  It therefore
separates backbone error from error introduced by the retrieval and adaptation
pipeline.

The main metrics are
\[
 \operatorname{MSE}=H^{-1}\lVert \hat y_0-y\rVert_2^2,
 \qquad
 \operatorname{nMSE}=H^{-1}\left\lVert
 \frac{\hat y_0-y}{\sigma(x)+\varepsilon}\right\rVert_2^2.
\]
Per-user aggregation and the worst 10\% user mean reveal whether a good global
mean hides systematic failures on a subset of series.

\goal{Establish the no-adaptation reference, measure its temporal and user-wise
heterogeneity, and provide the denominator against which every retrieval method
must be compared.}

\section{Windows and chronological protocol}
The integer $s$ is always the query date and the last observed date. The
look-back and target are
\[
 X_s=(s-L,s]=\{z_{s-L+1},\ldots,z_s\},\qquad
 Y_s=(s,s+H]=\{z_{s+1},\ldots,z_{s+H}\}.
\]
They contain exactly $L$ and $H$ values; the look-back never contains the
first forecast value. If a target interval begins at date $b$, its first query
is $s=b-1$. The look-back may cross that boundary, but the complete $Y_s$ must
lie inside the selected target interval. Consequently the held-out \Tthree
target dates do not shift when $L$ changes.

Raw dates are first split chronologically as
$(\Tzero,\Tone+\Ttwo,\Tthree)=(0.30,0.50,0.20)$. Extraction writes pooled
\code{adapt} (\Tone+\Ttwo) and held-out \code{eval} (\Tthree) payloads.
Each downstream model makes its own whole-query-date split of \code{adapt};
users sharing one query date can never be divided between \Tone and \Ttwo.
The control evaluates \Tthree. Each query contains all users at one date, but
the backbone is called as a univariate model with one channel per sample.

Dataset preparation automatically reads the sibling \code{config.json}.
Portable top-level options and adaptation-scoped options are combined before
the chronological split; user exclusions are additive and the selected config
path is logged. The same preparation contract is used by extraction, the
univariate control, and all downstream adapters.

The test profile uses Electricity 168--24 with sparse evaluation. The small
profile reproduces the three-dataset subset; full widens the Chronos control to
six datasets and adds the Cross-RAG setting. Ultra has the same direct Chronos
control because this reference launcher has no second backbone.

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
Component & Current sweep value \\
\midrule
Datasets & Traffic, Electricity, Solar \\
Look-back--horizon & 168--24, 504--24, 504--168, 504--504 \\
Backbone & Chronos \\
Input normalization & Instance normalization \\
Evaluation stride & 128 dates \\
Seed & 1 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Implementation:} \code{src/experiments/experiment\_univariate.py}\\
\textbf{Launcher:} \code{univariate.slurm}

It writes window-level losses,
a JSON summary, a tensor payload, error distributions, horizon-error curves,
user scatter plots, and prediction examples.  The primary comparison is the
same \Tthree MSE/nMSE later used by adapted methods.

\section{Interpretation}
This experiment is descriptive rather than trainable.  Any adapted method must
beat the control on the same queries and aggregation rule.  Improvements on
\Tone or \Ttwo alone are insufficient evidence because those periods are used
to fit adapters or gates.
\end{document}
