\documentclass[10pt]{article}
\input{common}
\title{Adaptation Experiment 2: Retrieval Baselines}
\author{}
\date{}
\begin{document}
\maketitle

\section{Signals and leakage-free retrieval}
For query date $s$, the exact window convention is
\[
 X_s=(s-L,s],\qquad Y_s=(s,s+H].
\]
Let $V=f(X_s)$ be the vanilla backbone forecast and let $C$ be the same
backbone conditioned on retrieved examples. For neighbor $j$, let $r_j$ be
its last look-back date, $Y_j$ its observed future, and $N_j=f(X_{r_j})$ its
vanilla forecast. All neighbor values are transferred to the query's level
and scale before use. Distance weights are
\[
 w_j=\frac{\exp(-\widetilde d_j)}{\sum_{\ell=1}^{k}
 \exp(-\widetilde d_\ell)},\qquad
 \widehat Y_h=\sum_{j=1}^{k}w_jY_{j,h}.
\]

In period-aligned retrieval, the last dates of the look-backs satisfy
$s-r_j\equiv0\pmod P$. In fixed mode every $(X_{r_j},Y_{r_j})$ lies wholly
inside \Tzero. In online mode a neighbor is eligible only when
$r_j+H\leq s$, so its complete future is observed by query time. A retrieved
future may overlap the query look-back, which is already observed, but it can
never overlap $Y_s$. These conditions prevent target leakage.

\goal{Compare all useful direct combinations of the available information
$V,C,Y_j,N_j$ before introducing a neural adapter.}

\section{Anchored models}
Every fitted forecast is written directly as $V+\text{linear signals}$.
There is no intercept and no design based only on differences such as
$Y_j-N_j$. The word ``residual'' below is only the agreed name for the model
containing both $Y_j$ and $N_j$.

A \emph{shared} model has one coefficient vector used at every horizon.
A \emph{horizon} model fits $H$ independent regressions. Each of those $H$
regressions has the same number of coefficients as the shared regression;
the total coefficient count is therefore multiplied by $H$.

\small
\begin{longtable}{p{.27\linewidth}p{.55\linewidth}r}
\toprule
Artifact name & Forecast at horizon $h$ & Coefficients \\
\midrule
\endhead
\code{vanilla} & $V_h$ & 0\\
\code{context\_forecast} & $C_h$ & 0\\
\code{aggr\_y} & $\widehat Y_h$ & 0\\
\code{y\_mean} & $k^{-1}\sum_jY_{j,h}$ & 0\\
\midrule
\code{aggr\_y\_mix\_shared}
  & $(1-\lambda)V_h+\lambda\widehat Y_h$ & 1\\
\code{aggr\_y\_mix\_horizon}
  & $(1-\lambda_h)V_h+\lambda_h\widehat Y_h$ & $1$ per $h$\\
\midrule
\code{context\_ridge\_shared}
  & $V_h+aV_h+bC_h$ & 2\\
\code{context\_ridge\_horizon}
  & $V_h+a_hV_h+b_hC_h$ & 2 per $h$\\
\code{aggr\_y\_ridge\_shared}
  & $V_h+aV_h+b\widehat Y_h$ & 2\\
\code{aggr\_y\_ridge\_horizon}
  & $V_h+a_hV_h+b_h\widehat Y_h$ & 2 per $h$\\
\code{y\_ridge\_shared}
  & $V_h+aV_h+\sum_jb_jY_{j,h}$ & $1+k$\\
\code{y\_ridge\_horizon}
  & $V_h+a_hV_h+\sum_jb_{j,h}Y_{j,h}$ & $1+k$ per $h$\\
\code{cov\_y\_ridge\_shared}
  & $V_h+aV_h+bC_h+\sum_jc_jY_{j,h}$ & $2+k$\\
\code{cov\_horizon\_ridge\_shared}
  & $V_h+aV_h+bC_h+c\widehat Y_h$ & 3\\
\code{cov\_horizon\_ridge\_horizon}
  & $V_h+a_hV_h+b_hC_h+c_h\widehat Y_h$ & 3 per $h$\\
\code{residual\_ridge\_shared}
  & $V_h+aV_h+\sum_jb_jY_{j,h}+\sum_jc_jN_{j,h}$ & $1+2k$\\
\code{residual\_ridge\_horizon}
  & $V_h+a_hV_h+\sum_jb_{j,h}Y_{j,h}+\sum_jc_{j,h}N_{j,h}$ & $1+2k$ per $h$\\
\code{full\_ridge\_shared}
  & $V_h+aV_h+bC_h+\sum_jc_jY_{j,h}+\sum_jd_jN_{j,h}$ & $2+2k$\\
\code{full\_ridge\_horizon}
  & $V_h+a_hV_h+b_hC_h+\sum_jc_{j,h}Y_{j,h}+\sum_jd_{j,h}N_{j,h}$ & $2+2k$ per $h$\\
\bottomrule
\end{longtable}
\normalsize

The two lambda models are the only retained convex mixtures. Their
least-squares estimate is projected onto $[0,1]$ with
$\lambda\leftarrow\min(1,\max(0,\lambda_{\rm LS}))$; the horizon variant
does this independently for each $\lambda_h$. Ridge coefficients are
unconstrained.

\section{Model-local split and validation protocol}
Extraction fixes only
\[
 \Tzero\;(30\%),\qquad \Tone+\Ttwo\;(50\%),\qquad \Tthree\;(20\%),
\]
and writes \code{adapt} and \code{eval} payloads. For these baselines,
\code{adapt} is split chronologically by whole query dates, with its final
20\% assigned to \Ttwo by default.

\begin{enumerate}
\item Accumulate exact float64 sufficient statistics on \Tone.
\item For every ridge family, fit every
$\alpha\in\{0,10^{-6},10^{-5},10^{-4},10^{-3},10^{-2},10^{-1},1,10\}$ and
select the value with lowest \Ttwo nMSE.
\item Refit that family on all of \Tone+\Ttwo using the selected $\alpha$.
\item Fit the lambda models directly on \Tone+\Ttwo because they have no
regularization hyperparameter to validate.
\item Score every selected model once on the complete untouched \Tthree.
\end{enumerate}

Features are RMS-standardized without centering, so a zero coefficient still
means no added signal. Chunked sufficient statistics change memory usage, not
the fitted objective. Optional fit-only sample caps never subsample final
\Tthree scoring. Optimistic \code{eval\_fit} variants remain available behind
an explicit flag for appendix diagnostics and are disabled by default.

\textbf{Code:} \code{src/experiments/extraction.py};
\code{src/adaptors/baselines/evaluate.py}.\\
\textbf{Launchers:} \code{extraction.slurm}; \code{baselines.slurm}.

\section{Interpretation}
\code{aggr\_y} tests analog forecasting directly; \code{context} tests whether
the backbone can consume retrieved context. The individual-$Y$ models test
whether learned neighbor-specific weights beat the fixed distance aggregation.
\code{residual} tests whether neighbor forecasts add information about neighbor
reliability, and \code{full} is the complete $V,C,Y_j,N_j$ model. All claims
must be based on the same complete \Tthree queries and compared with $V$.
\end{document}
