\documentclass[10pt]{article}
\input{common}
\title{Adaptation Experiment 3: Learned Retrieval Gates}
\author{}
\date{}
\begin{document}
\maketitle

\section{Candidates and targets}
Two gate families are evaluated. A context gate chooses between the vanilla
forecast $V$ and the context-informed forecast $C$. An aggregated-neighbor
gate chooses between $V$ and
\[
 \widehat Y_h=\sum_{j=1}^{k}w_jY_{j,h}.
\]
Write either candidate as $A\in\{C,\widehat Y\}$. The horizon-level advantage
used for supervision is
\[
 \Delta_h=(Y_h-V_h)^2-(Y_h-A_h)^2.
\]
A positive value means the candidate is better. Shared gates make one decision
for the complete forecast window using $\overline\Delta=H^{-1}\sum_h\Delta_h$;
horizon gates make $H$ separate decisions.

\goal{Learn when context or retrieved outcomes should replace the reliable
vanilla forecast, and compare conditional gates with no-feature and
target-aware upper bounds.}

\section{Engineered features}
Let
\[
 \mu_w(z)=\sum_jw_jz_j,\qquad
 \sigma_w(z)=\sqrt{\sum_jw_j(z_j-\mu_w(z))^2}.
\]
For each neighbor define the across-horizon summaries
\[
\begin{aligned}
 D^Y_j&=H^{-1}\sum_h(Y_{j,h}-V_h),
 &S^Y_j&=\operatorname{sd}_h(Y_{j,h}-V_h),\\
 D^N_j&=H^{-1}\sum_h(Y_{j,h}-N_{j,h}),
 &S^N_j&=\operatorname{sd}_h(Y_{j,h}-N_{j,h}).
\end{aligned}
\]
Both gate shapes receive
\[
 \mu_w(D^Y),\ \sigma_w(D^Y),\ \mu_w(S^Y),\qquad
 \mu_w(D^N),\ \sigma_w(D^N),\ \mu_w(S^N).
\]
They also receive the query mean and standard deviation; the mean and
between-neighbor standard deviation of neighbor look-back means and
look-back standard deviations; same-user ratio; mean neighbor age; weight
standard deviation and maximum; and mean retrieval distance.

The candidate features differ by gate shape:
\begin{itemize}
\item A shared gate receives
$\operatorname{mean}_h(A_h-V_h)$ and
$\operatorname{sd}_h(A_h-V_h)$.
\item The model for horizon $h$ receives only $A_h-V_h$, not the complete
vector $(A_1-V_1,\ldots,A_H-V_H)$. It additionally receives the local
$\mu_w(Y_{j,h}-V_h)$, $\sigma_w(Y_{j,h}-V_h)$,
$\mu_w(Y_{j,h}-N_{j,h})$, and $\sigma_w(Y_{j,h}-N_{j,h})$.
\end{itemize}
Thus every horizon gate has the same feature count and there are $H$
independent models; information from another candidate horizon cannot leak
through the candidate-difference feature.

\section{Models and artifact names}
\small
\begin{tabularx}{\linewidth}{p{.34\linewidth}X}
\toprule
Name pattern & Meaning \\
\midrule
\code{bayes\_<A>\_shared/horizon}
 & No-feature decision from the average \Tone+\Ttwo advantage.\\
\code{catboost\_<A>\_classifier\_}\\[-.25ex]
\code{shared/horizon}
 & CatBoost estimate of $\mathbb 1[\Delta>0]$; choose $A$ above probability
 $1/2$.\\
\code{catboost\_<A>\_regressor\_}\\[-.25ex]
\code{shared/horizon}
 & CatBoost estimate of the signed loss advantage; choose $A$ above zero.\\
\code{oracle\_<A>\_shared/horizon}
 & Target-aware upper bound evaluated on the scored split; never deployable.\\
\bottomrule
\end{tabularx}
\normalsize
Here \code{<A>} is \code{context} or \code{aggr\_y}.

\section{Validation protocol}
Extraction provides pooled \Tone+\Ttwo and fixed \Tthree payloads. For these
gates, the last 20\% of pooled query dates is \Ttwo by default:
\begin{enumerate}
\item fit CatBoost on \Tone with at most 300 trees, learning rate 0.03,
depth 4, and balanced classifier classes;
\item use \Ttwo only for early stopping and selection of the tree count;
\item instantiate a fresh model with that selected count and refit on all
of \Tone+\Ttwo;
\item score once on complete \Tthree.
\end{enumerate}
CatBoost does not require a separate hyperparameter search here: depth and
learning rate are fixed experimental choices, while \Ttwo selects effective
capacity through early stopping. Feature importance is saved separately for
each candidate, objective, and decision shape.

This refit is valid because $C$ and $\widehat Y$ are fixed candidates. For a
future gate whose candidate is itself trained, the stage-two protocol is
different: fit the candidate on \Tone and the gate on \Ttwo without refitting
the candidate afterward. Refitting the candidate on \Tone+\Ttwo would change
the gate inputs and make its \Ttwo supervision in-sample. Avoiding that bias
requires out-of-fold candidate predictions or an additional split.

\textbf{Launcher:} \code{gates.slurm}\\
\textbf{Module:} \code{src.adaptors.baselines.evaluate --family gates}

\section{Interpretation}
The no-feature gate measures whether one candidate wins globally. CatBoost
adds conditional decisions. The oracle gap is headroom, not performance.
Context and aggregated-neighbor gates answer different questions and should
both be reported against $V$ on identical \Tthree windows.
\end{document}
