\documentclass[10pt]{article}
\input{common}
\title{Adaptation Experiment 4: TS-IFA}
\author{}
\date{}
\begin{document}
\maketitle

\section{Theoretical overview and goal}
TS-IFA is a trainable adapter around a frozen forecast backbone.  It forms four
candidates: vanilla prediction $\hat y_0$, context-conditioned prediction
$\hat y_c$, a residual-attention prediction $\hat y_r$, and a memory-attention
prediction $\hat y_m$.  Cross-attention reads retrieved examples and the final
forecast is a horizon-wise convex mixture
\[
 \hat y_h=\sum_{b\in\{0,c,r,m\}}\pi_{b,h}\hat y_{b,h},
 \qquad \pi_{b,h}=\operatorname{softmax}_b(a_{b,h}).
\]
Residual attention uses neighbor backbone errors as values; memory attention
uses neighbor futures.  A third attention block produces mixture logits from
query and neighbor look-backs.  The horizon-specific four-way mixture is the
highest-capacity final component and is the first target for simplification if
\Ttwo overfitting persists.

\goal{Learn a small, backbone-agnostic adaptation mechanism that can exploit
retrieved forecasts, outcomes, and errors while remaining anchored to the
reliable vanilla forecast.}

\section{Objective and initialization}
With nMSE scaling, training minimizes
\[
 \mathcal L=\mathcal L_{\rm pred}
 +\beta\lVert\hat y-\hat y_0\rVert^2_{\rm n}
 +\gamma\bigl(\lVert\Delta_r-(y-\hat y_0)\rVert^2_{\rm n}
 +\lVert\Delta_m-(y-\hat y_0)\rVert^2_{\rm n}\bigr).
\]
Correction heads are zero-initialized.  Non-vanilla mixture logits start at
$-6$, so the initial adapter is close to the backbone and learns deviations
only when supported by data.

\section{Implementation details}
Each attention block uses four heads of dimension 32, followed by a 128-unit
GELU MLP.  Inputs are instance-normalized on the query scale.  AdamW uses
learning rate $10^{-5}$, weight decay $10^{-4}$, gradient clipping 1,
$\beta=\gamma=10^{-2}$, batch size 256, and 10,000 epochs.  Every epoch draws
one random batch from \Tone.  Full deterministic \Ttwo validation and interval
logging occur every 1,000 optimizer steps; \Tthree is evaluated only after
training.

The trainer can restore the lowest-nMSE \Ttwo checkpoint before touching
\Tthree and can stop after a configurable number of non-improving validation
evaluations.  The launcher enables best-checkpoint restoration but leaves early
stopping disabled by default for backward-compatible optimization length.
Dropout and reduced attention/MLP dimensions should be screened on one pilot
configuration before widening the sweep.  A scalar/shared or low-rank final
mixture is the next architecture ablation, not yet a default change.

The small sweep covers Electricity/Solar, 168--24 and 672--168, raw/instance
Euclidean retrieval, $k\in\{1,3,10\}$, and Chronos. The large profile adds
TabPFN-TS after the Chronos architecture is stable. Implementation is under
\code{src/adaptors/ts\_ifa/}; the launcher is
\code{ts\_ifa.slurm}. Saved artifacts include the model,
resolved configuration, full step history, interval-average train and \Ttwo
validation plot, \Tthree metrics, and branch predictions.

\section{Interpretation}
Report TS-IFA beside vanilla, context, residual, and memory branches.  This
distinguishes gains from better candidates from gains due to mixture learning.
Selection must use \Ttwo; the untouched \Tthree period is the only final test.
Attention weights are diagnostics and should not be interpreted as causal
feature importance.
\end{document}
